Search Strategies for Functional Logic Programming

### 2012-01-18 papers/atps2012/paper.tex
Final changes to ATPS version

\usepackage{graphicx}
\usepackage{url}
\usepackage{listings}
\lstset{aboveskip=1.2ex,
        belowskip=1.2ex,
        xleftmargin=2ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        basewidth=0.52em,
        basicstyle=\small\ttfamily}
\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
       }
\lstnewenvironment{curry}{}{}
\newcommand{\listline}{\vrule width0pt depth1.75ex}
\newcommand{\code}[1]{\texttt{\small{}#1}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92} % backslash
\newcommand{\us}{\char95} % underscore

@@

Another complete search strategy with a superior memory behavior
is iterative-deepening search.
Basically, it is depth-first search strategy with a depth-bound
which is incremented in each iteration.
Thus, we compute in each iteration a list of values
together with some information whether we have aborted
(due to the depth-bound) the computation of further possible values.
For this purpose, we define list structures that can also end
with an \code{Abort}:
%
\begin{curry}
data AbortList a = Nil | Cons a (AbortList a) | Abort
\end{curry}
%
and define the concatenation on such lists:
%
\begin{curry}
concA :: AbortList a -> AbortList a -> AbortList a
concA Abort       Abort       = Abort
concA Abort       Nil         = Abort
concA Abort       (Cons x xs) = Cons x (concA Abort xs)
concA Nil         ys          = ys
concA (Cons x xs) ys          = Cons x (concA xs ys)
\end{curry}
%
Now we define an operation to collect values in a search tree
within some level bounds (to avoid the repeated collection
of values found in each iteration):
%
\begin{curry}
collectInBounds :: Int -> Int -> SearchTree a -> AbortList a
collectInBounds oldbound newbound st = collectLevel newbound st
 where
  collectLevel _ Fail      = Nil
  collectLevel d (Value x) = if d<=newbound-oldbound then Cons x Nil
                                                     else Nil
  collectLevel d (Or x y)  =
    if d>0 then concA (collectLevel (d-1) x) (collectLevel (d-1) y)
           else Abort
\end{curry}
%
Now, the entire search strategy consists of
repeated calls to \code{collectInBounds} as long as the result list
is aborted. In order to experiment with different parameters,
the initial depth bound and the method to increase the depth bound
in each iteration are passed as parameters to the main operation:
%
\begin{curry}
allValuesIDS :: Int -> (Int -> Int) -> SearchTree a -> [a]
allValuesIDS initdepth incrdepth st =
  iterIDS initdepth (collectInBounds 0 initdepth st)
 where
  iterIDS _ Nil = []
  iterIDS n (Cons x xs) = x : iterIDS n xs
  iterIDS n Abort = let newdepth = incrdepth n
                     in iterIDS newdepth (collectInBounds n newdepth st)
\end{curry}
%
The key advantage of depth-first search
in comparison to breadth-first search is its memory behavior:
whereas breadth-first search has to store all nodes
in a level of the search tree in parallel, depth-first search
only needs to store the nodes in the branch from the root
to the current node under investigation.
Since iterative deepening uses depth-first search in each iteration,
it should have a memory behavior similarly to depth-first search
and, in case of wide search trees,
better than breadth-first search.
The price for this behavior is the recomputation of the initial
goal in each iteration. 
However, since we defined iterative deepening on a search tree,
recomputation is not required. Instead, the
already evaluated part of the search tree is kept in memory.
Therefore, we also implemented the iterative deepening strategy
for top-level goals without an explicit search tree
but with a recomputation of the initial goal in each iteration
(see Section~\ref{sec:benchmarks}).

@@

In this section we evaluate the practical behavior
of the various search strategies discussed so far.
Since they are only supported by the Curry implementation KiCS2,
we use only this system for our evaluation.
A general comparison of KiCS2 and other Curry implementations
can be found in \cite{BrasselHanusPeemoellerReck11}.
Since KiCS2 compiles Curry programs into Haskell programs,
we used the Glasgow Haskell Compiler
(GHC 7.0.4, option -O2) to compile and execute
the generated target programs.
All benchmarks were executed on a 32bit Linux machine
running Ubuntu ???? with an Intel Core 2 Duo (???GHz) processor and 4 GiB RAM.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark.
Due to the lack of precise measurements of the space behavior,
we measured only the execution times.

>>>>

\usepackage{graphicx}
\usepackage{url}
\usepackage{listings}
\lstset{aboveskip=1.2ex,
        belowskip=1.2ex,
        xleftmargin=2ex,
        showstringspaces=false, % no special string space
        mathescape=true,
        flexiblecolumns=false,
        basewidth=0.6em,
        basicstyle=\small\ttfamily}
\lstset{literate={->}{{$\rightarrow{}\!\!\!$}}3
       }
\lstnewenvironment{curry}{}{}
\newcommand{\listline}{\vrule width0pt depth1.75ex}
\newcommand{\code}[1]{\texttt{\small{}#1}}
\newcommand{\ccode}[1]{``\code{#1}''}
\newcommand{\bs}{\char92} % backslash
\newcommand{\us}{\char95} % underscore

@@

Another complete search strategy with a superior memory behavior
is iterative-deepening search.
Basically, it is depth-first search strategy with a depth-bound
which is incremented in each iteration.
Thus, we compute in each iteration a list of values
together with some information whether we have aborted
(due to the depth-bound) the computation of further possible values.
For this purpose, we define list structures that can also end
with an \code{Abort}:
%
\begin{curry}
data AbortList a = Nil | Cons a (AbortList a) | Abort
\end{curry}
%
and define the concatenation on such lists:
%
\begin{curry}
concA :: AbortList a -> AbortList a -> AbortList a
concA Abort       Abort       = Abort
concA Abort       Nil         = Abort
concA Abort       (Cons x xs) = Cons x (concA Abort xs)
concA Nil         ys          = ys
concA (Cons x xs) ys          = Cons x (concA xs ys)
\end{curry}
%
Now we define an operation to collect values in a search tree
within some level bounds (to avoid the repeated collection
of values found in each iteration):
%
\begin{curry}
collectInBounds :: Int -> Int -> SearchTree a -> AbortList a
collectInBounds oldbound newbound st = collectLevel newbound st
 where
  collectLevel _ Fail      = Nil
  collectLevel d (Value x) = if d <= newbound - oldbound
                             then Cons x Nil
                             else Nil
  collectLevel d (Or x y)  =
    if d>0
    then concA (collectLevel (d-1) x) (collectLevel (d-1) y)
    else Abort
\end{curry}
%
Now, the entire search strategy consists of
repeated calls to \code{collectInBounds} as long as the result list
is aborted. In order to experiment with different parameters,
the initial depth bound and the method to increase the depth bound
in each iteration are passed as parameters to the main operation:
%
\begin{curry}
allValuesIDS :: Int -> (Int -> Int) -> SearchTree a -> [a]
allValuesIDS initdepth incrdepth st =
  iterIDS initdepth (collectInBounds 0 initdepth st)
 where
  iterIDS _ Nil = []
  iterIDS n (Cons x xs) = x : iterIDS n xs
  iterIDS n Abort =
    let newdepth = incrdepth n
     in iterIDS newdepth (collectInBounds n newdepth st)
\end{curry}
%
The key advantage of depth-first search
in comparison to breadth-first search is its memory behavior:
whereas breadth-first search has to store all nodes
in a level of the search tree in parallel, depth-first search
only needs to store the nodes in the branch from the root
to the current node under investigation.
Since iterative deepening uses depth-first search in each iteration,
it should have a memory behavior similarly to depth-first search
and, in case of wide search trees,
better than breadth-first search.
The price for this behavior is the recomputation of the initial
goal in each iteration. 
However, since we defined iterative deepening on a search tree,
recomputation is not required. Instead, the
already evaluated part of the search tree is kept in memory.
Therefore, we also implemented the iterative deepening strategy
for top-level goals without an explicit search tree
but with a recomputation of the initial goal in each iteration
(see Section~\ref{sec:benchmarks}).

@@

In this section we evaluate the practical behavior
of the various search strategies discussed so far.
Since they are only supported by the Curry implementation KiCS2,
we use only this system for our evaluation.
A general comparison of KiCS2 and other Curry implementations
can be found in \cite{BrasselHanusPeemoellerReck11}.
Since KiCS2 compiles Curry programs into Haskell programs,
we used the Glasgow Haskell Compiler
(GHC 7.0.4, option -O2) to compile and execute
the generated target programs.
All benchmarks were executed on a 32bit Linux machine
running Ubuntu 11.10 with an Intel Core 2 Duo (2.13~GHz) processor and 4~GiB RAM.
The timings were performed with the time command measuring the
execution time (in seconds) of a compiled executable for each benchmark.
Due to the lack of precise measurements of the space behavior,
we measured only the execution times.
